Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-02-28 article-models-tt.tex
typos
getting the theory down to 60 rules

\begin{align*}
& \inference{\jfam{\Gamma}{A}}{\jfam{\Gamma}{A}} & & \inference{\jfameq{\Gamma}{A}{B}}{\jfameq{\Gamma}{A}{B}}\\
& \inference{\jfam{\Gamma}{A}\quad\jfameq{\Gamma}{A}{B}}{\jfam{\Gamma}{B}}
& & \inference{\jfam{\Gamma}{A}\quad\jfameq{\Gamma}{A}{B}}{\jfameq{\Gamma}{A}{B}}
\end{align*}

@@

\begin{align}
& \inference{}{\jctx{\emptyc}}\\
& \inference{\jctx{\Gamma}}{\jfam{\Gamma}{\emptyf\Gamma}}\\
& \inference{\jctx{\Gamma}}{\jterm{\Gamma}{\emptyf[\Gamma]}{\emptytm[\Gamma]}}\\
& \inference{\jterm{\Gamma}{\emptyf[\Gamma]}{x}}{\jtermeq{\Gamma}{\emptyf[\Gamma]}{x}{\emptytm[\Gamma]}}
\end{align}

@@

In the following inference rules we assume that $\jfam{\Gamma}{A}$.
\begin{align}
& \inference
  {\jfam{\ctxext{\Gamma}{B}}{Q}}
  {\jfameq{\ctxext{\Gamma}{A}}{\ctxwk{A}{\ctxext{B}{Q}}}{\ctxext{\ctxwk{A}{B}}{\ctxwk{A}{Q}}}}\\
& \inference
  {\jfam{\ctxext{{\Gamma}{B}}{Q}}{R}}
  {\jfameq
    {\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}
    {\ctxwk{A}{\ctxext{Q}{R}}}
    {\ctxext{\ctxwk{A}{Q}}{\ctxwk{A}{R}}}}
\end{align}

\subsubsection{Weakening is compatible with itself}\label{comp-ww}
We state judgmental equality rules expressing
that weakening is compatible with itself. These rules state that the following
diagram commutes given any two families $A$ and $B$ in context $\Gamma$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\Gamma}{\blank} \ar{r}{C\mapsto\ctxwk{B}{C}} \ar{d}[swap]{C\mapsto\ctxwk{A}{C}} & \jfam{\ctxext{\Gamma}{B}}{\blank} \ar{d}{Q\mapsto\ctxwk{A}{Q}}\\
\jfam{\ctxext{\Gamma}{A}}{\blank} \ar{r}[swap]{P\mapsto\ctxwk{{A}{B}}{P}} & \jfam{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\blank}
\end{tikzcd}
\end{equation*}
Thus, we get the following set of inference rules:
\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jfam{\Gamma}{C}}
          {\jfameq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{{B}{C}}}
            {\ctxwk{{A}{B}}{{A}{C}}}}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jterm{\Gamma}{C}{z}}
          {\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}
            {\ctxwk{A}{{B}{z}}}
            {\ctxwk{{A}{B}}{{A}{z}}}}\label{comp-ww-inf2}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jfam{\ctxext{\Gamma}{C}}{R}}
          {\jfameq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{A}{{B}{R}}}
            {\ctxwk{{A}{B}}{{A}{R}}}}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jterm{\Gamma}{R}{t}}
          {\jtermeq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{{A}{B}}{{A}{R}}}{\ctxwk{A}{{B}{t}}}{\ctxwk{{A}{B}}{{A}{t}}}}
            \label{comp-ww-inf4}
\end{align}
Note that strictly speaking, we did not need to require the inference in
\autoref{comp-ww-inf2} since this follows from the inference in \autoref{comp-ww-inf4}
and the fact that the weakening of the empty family is again the empty family.

@@

In the following inference rules we assume that
$\jfam{\Gamma}{A}$ and $\jfam{\ctxext{\Gamma}{A}}{P}$. 
\begin{align}
& \inference{\jfam{\Gamma}{B}}
    {\jfameq{\ctxext{{\Gamma}{A}}{P}}{\ctxwk{\ctxext{A}{P}}{B}}{\ctxwk{P}{{A}{B}}}}\\
& \inference{\jterm{\Gamma}{B}{y}}
    {\jtermeq{\ctxext{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}{\ctxwk{\ctxext{A}{P}}{y}}{\ctxwk{P}{{A}{y}}}}\\
& \inference{\jfam{\ctxext{\Gamma}{B}}{Q}}
    {\jfameq{\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{\ctxext{A}{P}}{Q}}{\ctxwk{P}{{A}{Q}}}}\\
& \inference{\jterm{\ctxext{\Gamma}{B}}{Q}{g}}
    {\jtermeq{\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{P}{{A}{Q}}}{\ctxwk{\ctxext{A}{P}}{g}}{\ctxwk{P}{{A}{g}}}}
\end{align}

@@

\subsubsection{Substitution is compatible with the empty context}
The fibers of the empty family are the empty families:
\begin{equation}
\inference{\jterm{\Gamma}{A}{x}}{\jfameq{\Gamma}{\subst{x}{\emptyf}}{\emptyf}}
\end{equation}

@@

\subsubsection{Substitution is compatible with extension}\label{comp-se}
Suppose $\jterm{\Gamma}{A}{x}$ in all of the following inference rules.
\begin{align}
& \inference{\jfam{\ctxext{{\Gamma}{A}}{P}}{Q}}
  {\jfameq{\Gamma}{\subst{x}{\ctxext{P}{Q}}}{\ctxext{\subst{x}{P}}{\subst{x}{Q}}}}\\
& \inference{\jfam{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}}
  {\jfameq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxext{Q}{R}}}{\ctxext{\subst{x}{Q}}{\subst{x}{R}}}}
\end{align}

\subsubsection{Substitution is compatible with weakening}\label{comp-sw}
The rules asserting the compatibility of substitution with weakening assert
that the following diagram commutes for any $\jterm{\Gamma}{A}{x}$ and any
$\jfam{\ctxext{\Gamma}{A}}{P}$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\ctxext{\Gamma}{A}}{\blank} \ar{d}[swap]{Q\mapsto\subst{x}{Q}} \ar{r}{Q\mapsto\ctxwk{P}{Q}} & \jfam{\ctxext{{\Gamma}{A}}{P}}{\blank} \ar{d}{R\mapsto\subst{x}{R}}\\ 
\jfam{\Gamma}{\blank} \ar{r}[swap]{B\mapsto\ctxwk{\subst{x}{P}}{B}} & \jfam{\ctxext{\Gamma}{\subst{x}{P}}}{\blank}
\end{tikzcd}
\end{equation*}
In the following inference rules we assume that $\jterm{\Gamma}{A}{x}$ and
$\jfam{\ctxext{\Gamma}{A}}{P}$:
\begin{align}
& \inference
    {\jfam{\ctxext{\Gamma}{A}}{Q}}
    {\jfameq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}{\ctxwk{\subst{x}{P}}{\subst{x}{Q}}}}\\
& \inference
    {\jfam{\ctxext{{\Gamma}{A}}{Q}}{R}}
    {\jfameq{\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}}{\subst{x}{\ctxwk{P}{R}}}{\ctxwk{\subst{x}{P}}{\subst{x}{R}}}}\\
& \inference
    {\jterm{\ctxext{\Gamma}{A}}{Q}{g}}
    {\jtermeq
      {\ctxext{\Gamma}{\subst{x}{P}}}
      {\subst{x}{\ctxwk{P}{Q}}}
      {\subst{x}{\ctxwk{P}{g}}}
      {\ctxwk{\subst{x}{P}}{\subst{x}{g}}}}\\
& \inference
    {\jterm{\ctxext{{\Gamma}{A}}{Q}}{R}{h}}
    {\jtermeq
      {\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}}
      {\subst{x}{\ctxwk{P}{R}}}
      {\subst{x}{\ctxwk{P}{h}}}
      {\ctxwk{\subst{x}{P}}{\subst{x}{h}}}}
\end{align}

@@

\begin{align}
&\inference{\jfam{\ctxext{{\Gamma}{A}}{P}}{Q}}
{\jfameq{\Gamma}{\subst{x}{{f}{Q}}}{\subst{{x}{f}}{{x}{Q}}}}\label{comp-ss-1}\\
&\inference{\jterm{\ctxext{{\Gamma}{A}}{P}}{Q}{g}}
{\jtermeq{\Gamma}{\subst{x}{{f}{Q}}}{\subst{x}{{f}{g}}}{\subst{{x}{f}}{{x}{g}}}}\label{comp-ss-2}\\
&\inference{\jfam{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}}
{\jfameq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{{x}{f}}{{x}{R}}}}\label{comp-ss-3}\\
&\inference{\jterm{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}{h}}
{\jtermeq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{x}{{f}{h}}}{\subst{{x}{f}}{{x}{h}}}}\label{comp-ss-4}
\end{align}

@@

\begin{align}
& \inference{\jfam{\ctxext{\Gamma}{B}}{Q}}{\jfameq{\ctxext{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{Q}}}}\\
& \inference{\jfam{\ctxext{{\Gamma}{B}}{Q}}{R}}{\jfameq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{Q}}}{\ctxwk{A}{\subst{y}{R}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{R}}}}\\
& \inference{\jterm{\ctxext{\Gamma}{B}}{Q}{g}}{\jtermeq{\ctxext{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}{\ctxwk{A}{\subst{y}{g}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{g}}}}\\
& \inference{\jterm{\ctxext{{\Gamma}{B}}{Q}}{R}{h}}{\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\ctxwk{A}{\subst{y}{h}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{h}}}}
\end{align}

>>>>

\begin{align*}
& \inference{\jtype{\Gamma}{A}}{\jfam{\Gamma}{A}} & & \inference{\jtypeeq{\Gamma}{A}{B}}{\jfameq{\Gamma}{A}{B}}\\
& \inference{\jtype{\Gamma}{A}\quad\jfameq{\Gamma}{A}{B}}{\jtype{\Gamma}{B}}
& & \inference{\jtype{\Gamma}{A}\quad\jfameq{\Gamma}{A}{B}}{\jtypeeq{\Gamma}{A}{B}}
\end{align*}

@@

\begin{align}
& \inference{}{\jctx{\emptyc}}\\
& \inference{\jctx{\Gamma}}{\jfam{\Gamma}{\emptyf[\Gamma]}}\\
& \inference{\jctx{\Gamma}}{\jterm{\Gamma}{\emptyf[\Gamma]}{\emptytm[\Gamma]}}\\
& \inference{\jterm{\Gamma}{\emptyf[\Gamma]}{x}}{\jtermeq{\Gamma}{\emptyf[\Gamma]}{x}{\emptytm[\Gamma]}}
\end{align}

@@

In the following inference rules we assume that $\jfam{\Gamma}{A}$.
\begin{align}
& \inference
  {\jfam{\ctxext{{\Gamma}{B}}{Q}}{R}}
  {\jfameq
    {\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}
    {\ctxwk{A}{\ctxext{Q}{R}}}
    {\ctxext{\ctxwk{A}{Q}}{\ctxwk{A}{R}}}}
\end{align}

\subsubsection{Weakening is compatible with itself}\label{comp-ww}
We state judgmental equality rules expressing
that weakening is compatible with itself. These rules state that the following
diagram commutes given any two families $A$ and $B$ in context $\Gamma$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\Gamma}{\blank} \ar{r}{C\mapsto\ctxwk{B}{C}} \ar{d}[swap]{C\mapsto\ctxwk{A}{C}} & \jfam{\ctxext{\Gamma}{B}}{\blank} \ar{d}{Q\mapsto\ctxwk{A}{Q}}\\
\jfam{\ctxext{\Gamma}{A}}{\blank} \ar{r}[swap]{P\mapsto\ctxwk{{A}{B}}{P}} & \jfam{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\blank}
\end{tikzcd}
\end{equation*}
Thus, we get the following set of inference rules:
\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jfam{\ctxext{\Gamma}{C}}{R}}
          {\jfameq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{A}{{B}{R}}}
            {\ctxwk{{A}{B}}{{A}{R}}}}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jterm{\ctxext{\Gamma}{C}}{R}{t}}
          {\jtermeq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{{A}{B}}{{A}{R}}}{\ctxwk{A}{{B}{t}}}{\ctxwk{{A}{B}}{{A}{t}}}}
            \label{comp-ww-inf4}
\end{align}

@@

In the following inference rules we assume that
$\jfam{\Gamma}{A}$ and $\jfam{\ctxext{\Gamma}{A}}{P}$. 
\begin{align}
& \inference{\jfam{\ctxext{\Gamma}{B}}{Q}}
    {\jfameq{\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{\ctxext{A}{P}}{Q}}{\ctxwk{P}{{A}{Q}}}}\\
& \inference{\jterm{\ctxext{\Gamma}{B}}{Q}{g}}
    {\jtermeq{\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{P}{{A}{Q}}}{\ctxwk{\ctxext{A}{P}}{g}}{\ctxwk{P}{{A}{g}}}}
\end{align}

@@

\subsubsection{Substitution is compatible with the empty context}
The fibers of the empty family are the empty families:
\begin{align}
& \inference{\jterm{\Gamma}{A}{x}}{\jfameq{\Gamma}{\subst{x}{\emptyf}}{\emptyf}}\\
& \inference{\jterm{\Gamma}{A}{x}\quad\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{\emptyf}}{\emptyf}}
\end{align}

@@

\subsubsection{Substitution is compatible with extension}\label{comp-se}
Suppose $\jterm{\Gamma}{A}{x}$ in all of the following inference rule.
\begin{align}
& \inference{\jfam{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}}
  {\jfameq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxext{Q}{R}}}{\ctxext{\subst{x}{Q}}{\subst{x}{R}}}}
\end{align}

\subsubsection{Substitution is compatible with weakening}\label{comp-sw}
The rules asserting the compatibility of substitution with weakening assert
that the following diagram commutes for any $\jterm{\Gamma}{A}{x}$ and any
$\jfam{\ctxext{\Gamma}{A}}{P}$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\ctxext{\Gamma}{A}}{\blank} \ar{d}[swap]{Q\mapsto\subst{x}{Q}} \ar{r}{Q\mapsto\ctxwk{P}{Q}} & \jfam{\ctxext{{\Gamma}{A}}{P}}{\blank} \ar{d}{R\mapsto\subst{x}{R}}\\ 
\jfam{\Gamma}{\blank} \ar{r}[swap]{B\mapsto\ctxwk{\subst{x}{P}}{B}} & \jfam{\ctxext{\Gamma}{\subst{x}{P}}}{\blank}
\end{tikzcd}
\end{equation*}
We plug in an extra layer of families to cover the most general case at once.
In the following inference rules we assume that $\jterm{\Gamma}{A}{x}$ and
$\jfam{\ctxext{\Gamma}{A}}{P}$:
\begin{align}
& \inference
  {\jfam{\ctxext{{\Gamma}{A}}{Q}}{R}}
  {\jfameq{\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}}{\subst{x}{\ctxwk{P}{R}}}{\ctxwk{\subst{x}{P}}{\subst{x}{R}}}}\\
& \inference
    {\jterm{\ctxext{{\Gamma}{A}}{Q}}{R}{h}}
    {\jtermeq
      {\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}}
      {\subst{x}{\ctxwk{P}{R}}}
      {\subst{x}{\ctxwk{P}{h}}}
      {\ctxwk{\subst{x}{P}}{\subst{x}{h}}}}
\end{align}

@@

\begin{align}
&\inference{\jfam{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}}
{\jfameq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{{x}{f}}{{x}{R}}}}\label{comp-ss-3}\\
&\inference{\jterm{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}{h}}
{\jtermeq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{x}{{f}{h}}}{\subst{{x}{f}}{{x}{h}}}}\label{comp-ss-4}
\end{align}

@@

\begin{align}
& \inference{\jfam{\ctxext{{\Gamma}{B}}{Q}}{R}}{\jfameq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{R}}}}\\
& \inference{\jterm{\ctxext{{\Gamma}{B}}{Q}}{R}{h}}{\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\ctxwk{A}{\subst{y}{h}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{h}}}}
\end{align}
